*lancer*: an R package for linearity assessment and visualisation of multiple curves

### 2023-01-21 Manuscript/manuscript.qmd
Swap statictical summary issue and plotting of many curves issue.

Linearity assessment plays a significant role in the validation of instrumentation and experimental procedures. Linearity can be tested by applying several graphical and numerical approaches. Commonly used spreadsheet software, e.g. Microsoft Excel, only allow the analyst to repetitively plot, view and analyse the linearity of curves one at a time, a tedious and time-consuming process. In addition, summary statistics of these curves are mostly limited to the Pearson Correlation Coefficient, which is insufficient to fully test the for linearity. While plotting of many curves at once and calculation of additional summary statistics for assessing linearity of curves can be done using programming languages like R, implementing this from scratch can be arduous task for the analyst. As such, we created the R package *lancer* so that curves can be analysed efficiently with a few functions. To our knowledge, this is currently not implemented in other software or programming languages. *lancer* furthermore reports the statistical results in tables/spreadsheets and records the plots in a pdf file. In addition, *lancer* can also create interactive trellis plots, displayed as a HTML folder, for exploratory analysis, helping analysts when dealing with large datasets.

@@

Linearity assessment is an important performance test of an analytical instrument or method when a linear response is assumed. It is applied in many fields in analytical sciences, such as assay development (@ross2003 and @hsieh2008), calibration/dilution studies (@rodríguez1993 and @sands2021) and laboratory tests. After analysis, curves representing response values versus concentration are plotted for each analyte of interest. As the accuracy of an analytical method is linked to linearity, a non-linear behaviour must be recognized and addressed accordingly. Visual inspection of curves is still useful but it must be accompanied by statistical tests as decision parameters.

To individually plot numerous curves and check for linearity is time-consuming. Furthermore, @sonnergaard2006 suggests that the Pearson Correlation Coefficient, a commonly used test, is an ineffective standalone numeric parameter to estimate linearity. While researchers like @vanloco2002, @sanchez2021 and @logue2018 have indicated other metrics for linearity evaluation, these metrics are generally not implemented in most software.

The R package *lancer* addresses these issues by assisting analysts to plot curves from many experiments, with additional metrics that better describe the characteristics of the curve. It also provides users with an interactive viewer to group, filter and sort the curves, allowing the examination of problematic cases, such as the ones generated by saturated signals.

>>>>

Linearity assessment plays a significant role in the validation of instrumentation and experimental procedures. Linearity can be tested by applying several graphical and numerical approaches. Summary statistics of these curves are mostly limited to the Pearson Correlation Coefficient, which is insufficient to fully test the for linearity. In addition, spreadsheet software, e.g. Microsoft Excel, only allow the analyst to repetitively plot, view and analyse the linearity of curves one at a time, a tedious and time-consuming process. While plotting of many curves at once and calculation of additional summary statistics for assessing linearity of curves can be done using programming languages like R, implementing this from scratch can be arduous task for the analyst. As such, we created the R package *lancer* so that curves can be analysed efficiently with a few functions. To our knowledge, this is currently not implemented in other software or programming languages. *lancer* furthermore reports the statistical results in tables/spreadsheets and records the plots in a pdf file. In addition, *lancer* can also create interactive trellis plots, displayed as a HTML folder, for exploratory analysis, helping analysts when dealing with large datasets.

@@

Linearity assessment is an important performance test of an analytical instrument or method when a linear response is assumed. It is applied in many fields in analytical sciences, such as assay development (@ross2003 and @hsieh2008), calibration/dilution studies (@rodríguez1993 and @sands2021) and laboratory tests. After analysis, curves representing response values versus concentration are plotted for each analyte of interest. As the accuracy of an analytical method is linked to linearity, a non-linear behaviour must be recognized and addressed accordingly.

Visual inspection of curves is useful but it must be accompanied by statistical tests as decision parameters. A commonly used test is the Pearson Correlation Coefficient. However, @sonnergaard2006 suggests that it is an ineffective standalone numeric parameter to estimate linearity. While researchers like @vanloco2002, @sanchez2021 and @logue2018 have indicated other metrics for linearity evaluation, these metrics are generally not implemented in most software. Furthermore, using spreadsheet software, e.g. Microsoft Excel, to individually plot numerous curves is time-consuming. While programming languages like R can plot numerous curve easily, it is hard to implement for a novice analyst.

The R package *lancer* addresses these issues by assisting analysts useful functions to plot many curves with additional metrics that better describe the curves' characteristics. It also provides analysts with an interactive viewer to group, filter and sort the curves, for examination of problematic cases, such as curves generated by saturated signals.
